Operator evolution from the similarity renormalization group and the Magnus expansion

### 2019-10-28 Notes/Operator_evolution/operator_evolution.tex
Update operator_evolution.tex
Added a bullet point to operator evolution section.
Updated operator_evolution notes
Minor changes.

\noindent{%
\textbf{Modern nuclear potentials and universality.}
}
\\
There has been tremendous progress in development of NN interactions from chiral effective field theory ($\chi^{EFT}$) in the last decade. Previous SRG studies were limited to phenomenological interactions or some of the first non-local chiral interactions \cite{Entem:2003ft}, but with much more variance in modern chiral interactions, we would like to revisit the question of how different potentials and other operators change under SRG transformations. Since $\chi^{EFT}$ is an effective field theory, different potentials will still give the same $S$-matrix, that is, the potentials are phase equivalent up to some higher energy. However, the potentials themselves can differ in many ways such as the regulator functions, cutoff, order, and so forth, and thus, the potential matrix elements are not the same. Dainton \textit{et al.} \cite{Dainton:2013axa} demonstrated SRG transformations drive different NN potentials to the same low-momentum matrix elements, that is, a flow to universality up to the momentum value of phase inequivalence. We address the question of universality for modern chiral potentials as well as how other operators change with respect to the corresponding SRG transformations.
\\

@@

%	\subfloat[]{%
%	\includegraphics[clip,width=0.9\columnwidth]{SRG_potentials/potential_contours_kvnn111_1P1_Wegner_channel_label}%
%	}
%	\caption{Matrix elements of the RKE N$^4$LO potential SRG-evolving in $\lambda$ right to left under transformations with the Wegner generator in the $^1$S$_0$ (a), $^3$S$_1$ (b), and $^1$P$_1$ (c) channels.}
%	\label{fig:potential_contours_RKE_Wegner}
%\end{figure}
%
\noindent{%
-- In Fig.~\ref{fig:potential_contours_3S1_Wegner} we consider three different SRG-evolved potentials in the $^3$S$_1$ channel: EM N$^3$LO (500 MeV cutoff) \cite{Entem:2003ft}, RKE N$^4$LO (450 MeV cutoff) \cite{Reinert:2017usi}, and Gezerlis et al. N$^2$LO (1 fm cutoff) \cite{Gezerlis:2014zia}. The major difference in these three potentials are the regulator functions in the contact and pion-exchange terms. The EM N$^3$LO interaction is a non-local potential where both contact and pion-exchange interactions feature a non-local regulator function of the form exp$[-(p/\Lambda)^{2n}-(p'/\Lambda)^{2n}]$, where $\Lambda$ is the momentum-space cutoff and $n$ is an integer. However, a non-local regulator function for pion-exchange contributions can introduce regulator artifacts for cutoffs $\Lambda$ lower than the breakdown scale $\Lambda_b$. Several semi-local chiral potentials have been introduced to reduce regulator artifacts, such as the RKE N$^4$LO potential. Here, a local regulator function is applied for the long-range interactions in momentum-space, while a non-local regulator function is used for the short-range interactions. In some instances, non-local interactions are not suitable for $\textit{ab initio}$ approaches such as the quantum Monte Carlo (QMC) method motivating the need for fully local potentials. The Gezerlis et al. N$^2$LO potential is an example of a local interaction where the long-range terms have a local regulator function in coordinate-space and the short-range terms have a local regulator function in momentum-space.
}
\\
-- Takeaways from Fig.~\ref{fig:potential_contours_3S1_Wegner}: completely different at $\lambda=6$ fm$^{-1}$ but low-momentum matrix elements are similar at $\lambda=1$ fm$^{-1}$. Decoupled, low-momentum matrix elements are necessarily the same since the pion-exchange terms are calculated explicitly. (Cutoff dependence can play a role though for lower cutoffs.)
%
\begin{figure}[H]
	\centering
	\subfloat[]{%
	\includegraphics[clip,width=0.9\columnwidth]{SRG_potentials/potential_contours_kvnn10_3S1_Wegner_potential_label}%
	}

@@

	\subfloat[]{%
	\includegraphics[clip,width=0.9\columnwidth]{SRG_potentials/potential_contours_kvnn222_3S1_Wegner_potential_label}%
	}
	\caption{Matrix elements of the EM N$^3$LO (a), RKE N$^4$LO (b), and Gezerlis et al. N$^2$LO (c) potentials SRG-evolving in $\lambda$ right to left under transformations with the Wegner generator in the $^3$S$_1$ channel.}
	\label{fig:potential_contours_3S1_Wegner}
\end{figure}
%
\noindent{%
-- Fig.~\ref{fig:potential_contours_1P1_RKE} shows the SRG-evolved RKE N$^4$LO (450 MeV cutoff) potential in the $^1$P$_1$ channel for two SRG generators: the Wegner and block-diagonal generators which drive the potential to band- and block-diagonal form, respectively. We continue to evolve to band-diagonal form with respect to the parameter $\lambda$, but for the block-diagonal generator, we label sub-plots with the parameter $\Lambda$ which characterizes the sharp cutoff in decoupling the low- and high-momentum matrix elements.
}
\\
-- Takeaways from Fig.~\ref{fig:potential_contours_1P1_RKE}: smooth decoupling for Wegner and sharp for block-diagonal, each unique generator should have its own type of universality. Check this more quantitatively by comparing matrix element ``slices''.
%
\begin{figure}[H]
	\centering
	\subfloat[]{%
	\includegraphics[clip,width=0.9\columnwidth]{SRG_potentials/potential_contours_kvnn111_1P1_Wegner_generator_label}%
	}

	\subfloat[]{%
	\includegraphics[clip,width=0.9\columnwidth]{SRG_potentials/potential_contours_kvnn111_1P1_Block-diag_generator_label}%
	}
	\caption{Matrix elements of the RKE N$^4$LO potential SRG-evolving right to left under transformations with Wegner (a) and block-diagonal (b) generators in the $^1$P$_1$ channel. Here, we use $\lambda$ for Wegner evolution in the top row and $\Lambda$ for block-diagonal evolution in the bottom row. For block-diagonal evolution, we fix $\lambda=1.5$ fm$^{-1}$.}
	\label{fig:potential_contours_1P1_RKE}
\end{figure}
%
\noindent{%
-- Fig.~\ref{fig:phase_shifts} shows the NN phase shifts of EM N$^3$LO, RKE N$^4$LO, and Gezerlis et al. N$^2$LO potentials in the $^1$S$_0$, $^3$S$_1$, and $^1$P$_1$ channels. In \cite{Dainton:2013axa}, it was found that phase equivalence up to some value of momentum $k$ implies matrix element equivalence up to the same value of $k$ in SRG-evolved potentials. We verify this conclusion by checking the matrix elements in Fig.~\ref{fig:potential_slices_1S0} where we see a collapse of the different potential matrix elements to the same line in the last column ($\lambda=1$ fm$^{-1}$). Note, Figs.~\ref{fig:potential_slices_1S0}-\ref{fig:potential_slices_1P1} show both Wegner and block-diagonal evolution where the solid lines correspond to the Wegner generator and dash-dotted to block-diagonal evolution. We see each generator collapses the potential to a different form because the low- and high-momentum matrix elements decouple in a different manner.
}
\\
-- In the $^3$S$_1$ channel, we see a slight deviation in the lowest momentum potential matrix element from EM N$^3$LO and the other two potentials. This is due to a minor difference in the deuteron binding energy ($\approx 1\%$ difference).
\\
-- The bottom row of Fig.~\ref{fig:potential_slices_1P1} shows $V(k,0.5)$ instead of $V(k,0)$ because the far off-diagonal matrix elements in the $^1$P$_1$ channel are all zero.
\begin{figure}[H]
	\centering
	\subfloat[]{%
	\includegraphics[clip,width=0.3\columnwidth]{SRG_observables/phase_shifts_1S0_kvnns_10_111_222}%
	}
	\quad
	\subfloat[]{%
	\includegraphics[clip,width=0.3\columnwidth]{SRG_observables/phase_shifts_3S1_kvnns_10_111_222}%
	}
	\quad
	\subfloat[]{%
	\includegraphics[clip,width=0.3\columnwidth]{SRG_observables/phase_shifts_1P1_kvnns_10_111_222}%
	}
	\caption{$^1$S$_0$ (a), $^3$S$_1$ (b), and $^1$P$_1$ (c) phase shifts for the EM N$^3$LO (solid black), RKE N$^4$LO (red dash-dotted), and Gezerlis et al. N$^2$LO (blue dashed) potentials.}
	\label{fig:phase_shifts}
\end{figure}
%
\begin{figure}[H]
	\centering
	\subfloat[]{%
	\includegraphics[clip,width=0.9\columnwidth]{SRG_potentials/potential_diag_1S0_kvnns_10_111_222_lamb1,0}%
	}

	\subfloat[]{%
	\includegraphics[clip,width=0.9\columnwidth]{SRG_potentials/potential_off-diag_1S0_kvnns_10_111_222_lamb1,0}%
	}
	\caption{Diagonal (a) and far off-diagonal (b) matrix elements of the EM N$^3$LO (black), RKE N$^4$LO (red) and Gezerlis et al. N$^2$LO (blue) potentials SRG-evolving right to left under transformations with Wegner (solid) and block-diagonal (dash-dotted) generators in the $^1$S$_0$ channel. Here, we use $\lambda$ for Wegner evolution and $\Lambda$ for block-diagonal evolution. For block-diagonal evolution, we fix $\lambda=1$ fm$^{-1}$.}
	\label{fig:potential_slices_1S0}
\end{figure}
%
\begin{figure}[H]
	\centering
	\subfloat[]{%
	\includegraphics[clip,width=0.9\columnwidth]{SRG_potentials/potential_diag_3S1_kvnns_10_111_222_lamb1,0}%
	}

@@

	\subfloat[]{%
	\includegraphics[clip,width=0.9\columnwidth]{SRG_operators/momentum_projection_off-diag_q3,00_3S1_kvnns_10_111_222_lamb1,5}%
	}
	\caption{Diagonal (a) and far off-diagonal (b) matrix elements of the momentum projection operator under SRG transformations with EM N$^3$LO (black), RKE N$^4$LO (red) and Gezerlis et al. N$^2$LO (blue) potentials evolving right to left with Wegner (solid) and block-diagonal (dash-dotted) generators in the $^3$S$_1$ channel. Here, we use $\lambda$ for Wegner evolution and $\Lambda$ for block-diagonal evolution. For block-diagonal evolution, we fix $\lambda=1.5$ fm$^{-1}$.}
	\label{fig:momentum_proj_3S1}
\end{figure}
%
-- Note, the unitary transformations and evolved wave functions for different potentials are not the same (at low momentum too) which implies that evolved operators cannot have the same universality property as the evolved potentials since they are dependent on the evolved wave functions.
\\
-- $\hat{r}^2$ operator.
\\
-- Relevant equations: operator in coordinate-space and Hankel transformation.

>>>>

\noindent{%
\textbf{Modern nuclear potentials and universality.}
}
\\
There has been tremendous progress in development of NN interactions from chiral effective field theory ($\chi^{EFT}$) in the last decade. Previous SRG studies were limited to phenomenological interactions or some of the first non-local chiral interactions \cite{Entem:2003ft}, but with much more variance in modern chiral interactions, we would like to revisit the question of how different potentials and other operators change under SRG transformations. Since $\chi^{EFT}$ is an effective field theory, different potentials will still give the same $S$-matrix, that is, the potentials are phase equivalent up to some higher energy. However, the potentials themselves can differ in many ways such as the regulator functions, cutoff, order, and so forth, and thus, the potential matrix elements are not the same. Dainton \textit{et al.}~\cite{Dainton:2013axa} demonstrated SRG transformations drive different NN potentials to the same low-momentum matrix elements, that is, a flow to universality up to the momentum value of phase inequivalence. We address the question of universality for modern chiral potentials as well as how other operators change with respect to the corresponding SRG transformations.
\\

@@

%	\subfloat[]{%
%	\includegraphics[clip,width=0.9\columnwidth]{SRG_potentials/potential_contours_kvnn111_1P1_Wegner_channel_label}%
%	}
%	\caption{Matrix elements of the RKE N$^4$LO potential SRG-evolving in $\lambda$ right to left under transformations with the Wegner generator in the $^1$S$_0$ (a), $^3$S$_1$ (b), and $^1$P$_1$ (c) channels.}
%	\label{fig:potential_contours_RKE_Wegner}
%\end{figure}
%
\noindent{%
-- In Fig.~\ref{fig:potential_contours_3S1_Wegner} we consider three different SRG-evolved potentials in the $^3$S$_1$ channel: EM N$^3$LO (500 MeV cutoff) \cite{Entem:2003ft}, RKE N$^4$LO (450 MeV cutoff) \cite{Reinert:2017usi}, and Gezerlis et al.~N$^2$LO (1 fm cutoff) \cite{Gezerlis:2014zia}. The major difference in these three potentials are the regulator functions in the contact and pion-exchange terms. The EM N$^3$LO interaction is a non-local potential where both contact and pion-exchange interactions feature a non-local regulator function of the form exp$[-(p/\Lambda)^{2n}-(p'/\Lambda)^{2n}]$, where $\Lambda$ is the momentum-space cutoff and $n$ is an integer. However, a non-local regulator function for pion-exchange contributions can introduce regulator artifacts for cutoffs $\Lambda$ lower than the breakdown scale $\Lambda_b$. Several semi-local chiral potentials have been introduced to reduce regulator artifacts, such as the RKE N$^4$LO potential. Here, a local regulator function is applied for the long-range interactions in momentum-space, while a non-local regulator function is used for the short-range interactions. In some instances, non-local interactions are not suitable for $\textit{ab initio}$ approaches such as the quantum Monte Carlo (QMC) method motivating the need for fully local potentials. The Gezerlis et al.~N$^2$LO potential is an example of a local interaction where the long-range terms have a local regulator function in coordinate-space and the short-range terms have a local regulator function in momentum-space.
}
\\
-- Takeaways from Fig.~\ref{fig:potential_contours_3S1_Wegner}: completely different at $\lambda=6$ fm$^{-1}$ but low-momentum matrix elements are similar at $\lambda=1$ fm$^{-1}$. Decoupled, low-momentum matrix elements are necessarily the same since the pion-exchange terms are calculated explicitly. (Cutoff dependence can play a role though for lower cutoffs.)
%
\begin{figure}[H]
	\centering
	\subfloat[]{%
	\includegraphics[clip,width=0.9\columnwidth]{SRG_potentials/potential_contours_kvnn10_3S1_Wegner_potential_label}%
	}

@@

	\subfloat[]{%
	\includegraphics[clip,width=0.9\columnwidth]{SRG_potentials/potential_contours_kvnn222_3S1_Wegner_potential_label}%
	}
	\caption{Matrix elements of the EM N$^3$LO (a), RKE N$^4$LO (b), and Gezerlis et al.~N$^2$LO (c) potentials SRG-evolving in $\lambda$ right to left under transformations with the Wegner generator in the $^3$S$_1$ channel.}
	\label{fig:potential_contours_3S1_Wegner}
\end{figure}
%
\noindent{%
-- Fig.~\ref{fig:potential_contours_1P1_RKE} shows the SRG-evolved RKE N$^4$LO (450 MeV cutoff) potential in the $^1$P$_1$ channel for two SRG generators: the Wegner and block-diagonal generators which drive the potential to band- and block-diagonal form, respectively. We continue to evolve to band-diagonal form with respect to the parameter $\lambda$, but for the block-diagonal generator, we label sub-plots with the parameter $\Lambda$ which characterizes the sharp cutoff in decoupling the low- and high-momentum matrix elements.
}
\\
-- Takeaways from Fig.~\ref{fig:potential_contours_1P1_RKE}: smooth decoupling for Wegner and sharp for block-diagonal, each unique generator should have its own type of universality. Check this more quantitatively by comparing matrix element ``slices''.
%
\begin{figure}[H]
	\centering
	\subfloat[]{%
	\includegraphics[clip,width=0.9\columnwidth]{SRG_potentials/potential_contours_kvnn111_1P1_Wegner_generator_label}%
	}

	\subfloat[]{%
	\includegraphics[clip,width=0.9\columnwidth]{SRG_potentials/potential_contours_kvnn111_1P1_Block-diag_generator_label}%
	}
	\caption{Matrix elements of the RKE N$^4$LO potential SRG-evolving right to left under transformations with Wegner (a) and block-diagonal (b) generators in the $^1$P$_1$ channel. Here, we use $\lambda$ for Wegner evolution in the top row and $\Lambda$ for block-diagonal evolution in the bottom row. For block-diagonal evolution, we fix $\lambda=1.5$ fm$^{-1}$.}
	\label{fig:potential_contours_1P1_RKE}
\end{figure}
%
\noindent{%
-- Fig.~\ref{fig:phase_shifts} shows the NN phase shifts of EM N$^3$LO, RKE N$^4$LO, and Gezerlis et al.~N$^2$LO potentials in the $^1$S$_0$, $^3$S$_1$, and $^1$P$_1$ channels. In \cite{Dainton:2013axa}, it was found that phase equivalence up to some value of momentum $k$ implies matrix element equivalence up to the same value of $k$ in SRG-evolved potentials. We verify this conclusion by checking the matrix elements in Fig.~\ref{fig:potential_slices_1S0} where we see a collapse of the different potential matrix elements to the same line in the last column ($\lambda=1$ fm$^{-1}$). Note, Figs.~\ref{fig:potential_slices_1S0}-\ref{fig:potential_slices_1P1} show both Wegner and block-diagonal evolution where the solid lines correspond to the Wegner generator and dash-dotted to block-diagonal evolution. We see each generator collapses the potential to a different form because the low- and high-momentum matrix elements decouple in a different manner.
}
\\
-- In the $^3$S$_1$ channel, we see a slight deviation in the lowest momentum potential matrix element from EM N$^3$LO and the other two potentials. This is due to a minor difference in the deuteron binding energy ($\approx 1\%$ difference).
\\
-- The bottom row of Fig.~\ref{fig:potential_slices_1P1} shows $V(k,0.5)$ instead of $V(k,0)$ because the far off-diagonal matrix elements in the $^1$P$_1$ channel are all zero.
\begin{figure}[H]
	\centering
	\subfloat[]{%
	\includegraphics[clip,width=0.3\columnwidth]{SRG_observables/phase_shifts_1S0_kvnns_10_111_222}%
	}
	\quad
	\subfloat[]{%
	\includegraphics[clip,width=0.3\columnwidth]{SRG_observables/phase_shifts_3S1_kvnns_10_111_222}%
	}
	\quad
	\subfloat[]{%
	\includegraphics[clip,width=0.3\columnwidth]{SRG_observables/phase_shifts_1P1_kvnns_10_111_222}%
	}
	\caption{$^1$S$_0$ (a), $^3$S$_1$ (b), and $^1$P$_1$ (c) phase shifts for the EM N$^3$LO (solid black), RKE N$^4$LO (red dash-dotted), and Gezerlis et al.~N$^2$LO (blue dashed) potentials.}
	\label{fig:phase_shifts}
\end{figure}
%
\begin{figure}[H]
	\centering
	\subfloat[]{%
	\includegraphics[clip,width=0.9\columnwidth]{SRG_potentials/potential_diag_1S0_kvnns_10_111_222_lamb1,0}%
	}

	\subfloat[]{%
	\includegraphics[clip,width=0.9\columnwidth]{SRG_potentials/potential_off-diag_1S0_kvnns_10_111_222_lamb1,0}%
	}
	\caption{Diagonal (a) and far off-diagonal (b) matrix elements of the EM N$^3$LO (black), RKE N$^4$LO (red) and Gezerlis et al.~N$^2$LO (blue) potentials SRG-evolving right to left under transformations with Wegner (solid) and block-diagonal (dash-dotted) generators in the $^1$S$_0$ channel. Here, we use $\lambda$ for Wegner evolution and $\Lambda$ for block-diagonal evolution. For block-diagonal evolution, we fix $\lambda=1$ fm$^{-1}$.}
	\label{fig:potential_slices_1S0}
\end{figure}
%
\begin{figure}[H]
	\centering
	\subfloat[]{%
	\includegraphics[clip,width=0.9\columnwidth]{SRG_potentials/potential_diag_3S1_kvnns_10_111_222_lamb1,0}%
	}

@@

	\subfloat[]{%
	\includegraphics[clip,width=0.9\columnwidth]{SRG_operators/momentum_projection_off-diag_q3,00_3S1_kvnns_10_111_222_lamb1,5}%
	}
	\caption{Diagonal (a) and far off-diagonal (b) matrix elements of the momentum projection operator under SRG transformations with EM N$^3$LO (black), RKE N$^4$LO (red) and Gezerlis et al.~N$^2$LO (blue) potentials evolving right to left with Wegner (solid) and block-diagonal (dash-dotted) generators in the $^3$S$_1$ channel. Here, we use $\lambda$ for Wegner evolution and $\Lambda$ for block-diagonal evolution. For block-diagonal evolution, we fix $\lambda=1.5$ fm$^{-1}$.}
	\label{fig:momentum_proj_3S1}
\end{figure}
%
-- Note, the unitary transformations and evolved wave functions for different potentials are not the same (at low momentum too) which implies that evolved operators cannot have the same universality property as the evolved potentials since they are dependent on the evolved wave functions.
\\
-- $\hat{r}^2$ operator.
\\
-- Relevant equations: operator in coordinate-space and Hankel transformation.
\\
-- Add continuum state momentum distributions and how to understand operator evolution. If
